<!--
%\VignetteIndexEntry{A C++ API for R matrix objects}
%\VignettePackage{BiocStyle}
%\VignetteEngine{knitr::knitr}
-->

```{r, echo=FALSE, results="hide", message=FALSE}
require(knitr)
opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

```{r style, echo=FALSE, results='asis'}
BiocStyle::markdown()
```

# A common C++ API for all types of R matrices
Package: `r Rpackage("beachmat")` <br />
Author: Aaron Lun (alun@wehi.edu.au) <br />
Compilation date: `r Sys.Date()`

# Introduction

The _beachmat_ package provides a C++ API for handling a variety of R matrix types.
The aim is to abstract the specific type of matrix when writing C++ extensions to R packages, thus simplifying the processing of data stored in matrices or matrix-like objects.
Currently, the API supports double-precision, integer and logical matrices.
Supported classes include base `matrix` objects, a number of classes from the `r CRANpkg("Matrix")` package, and disk-backed matrices from the `r CRANpkg("HDF5Array")` package.


# Linking to the package

## Prerequisites

The _beachmat_ package currently has several dependencies:

- The build system should support the C++11 standard.
- The system should have the HDF5 libraries installed (see https://support.hdfgroup.org/HDF5/).
Currently, we are not particularly smart about detecting non-default installation locations.
This should hopefully be rectified in the future.
- `r CRANpkg("Rcpp")` should be installed.

Note that most of the following instructions are ripped from the `r Biocpkg("Rhtslib")` vignette.

## Link to the library

To link successfully to the _beachmat_ library, a package must include *both* a `src/Makevars.win` *and* `src/Makevars` file.  

**Note**: the contents of `src/Makevars.win` and `src/Makevars` are almost identical, but not quite. 
Be careful of the differences.

Create a `src/Makevars.win` file with the following lines:

```
BEACHLIB_LIBS=$(shell echo 'beachmat::pkgconfig("PKG_LIBS")'|\
    "${R_HOME}/bin/R" --vanilla --slave)
PKG_LIBS=$(BEACHLIB_LIBS)
```

... and a `src/Makevars` file with the following lines:

```
BEACHLIB_LIBS=`echo 'beachmat::pkgconfig("PKG_LIBS")'|\
    "${R_HOME}/bin/R" --vanilla --slave`
PKG_LIBS=$(BEACHLIB_LIBS)
```

The statement for each platfrom modifies the `$PKG_LIBS` variable. 
If your package needs to add to the `$PKG_LIBS` variable, do so by adding to the `PKG_LIBS=$(RHTSLIB_LIBS)` line.
For example:

```
PKG_LIBS=$(RHTSLIB_LIBS) -L/path/to/foolib -lfoo
```

## Find headers

In order for the C/C++ compiler to find the _beachmat_ package headers during installation, add `beachmat` and `Rcpp` to the `LinkingTo` field of the `DESCRIPTION` file:

```
LinkingTo: Rcpp, beachmat
``` 

Note that as of R 3.0.2 `LinkingTo` values can include version specifications, e.g., `LinkingTo: Rhtslib (>= 0.99.10)`.
    
In C or C++ code files, use standard techniques, e.g., `#include "numeric_matrix.h"` (see below for more details). 
Header files are available for perusal at the following location (enter in an R session):

```{R headers}
system.file(package="beachmat", "include")
```

## Initializing `r CRANpkg("Rcpp")`

One final task is to ensure that `r CRANpkg("Rcpp")` is initialized when your package is loaded.
This requires addition of `Rcpp` to the `Imports` field of the `DESCRIPTION` file:

```
Imports: Rcpp
```

... and a corresponding  `importFrom` specification in the `NAMESPACE` file:

```
importFrom(Rcpp, sourceRcpp)
```
 
(The exact function to be imported doesn't matter, as long as the namespace is loaded.)
Check out the `r CRANpkg("Rcpp")` documentation for more details.

# Overview of the API

## An example with a numeric matrix

We demonstrate the use of the API for numeric matrices.
First, we include the relevant header file (in this case, `numeric_matrix.h`):

```
#include "numeric_matrix.h"
```

A double-precision matrix object `dmat` is handled in C++ by passing the `SEXP` struct from `.Call` to `create_numeric_matrix`:

```
std::shared_dptr<numeric_matrix> dptr=create_numeric_matrix(dmat);
```

This creates a shared pointer that points to an object of the `numeric_matrix` virtual class.
The exact class depends on the type of matrix in `dmat`, though the behaviour of the user-level functions are not affected by this detail.
(Note that this requires C++11 to work.)

The available methods for this object are:

- `dptr->get_nrow()` returns an integer number of rows (`nrow`).
- `dptr->get_ncol()` returns an integer number of columns (`ncol`).
- `dptr->get_row(r)` returns a constant pointer to an array of doubles of length `ncol`, containing values at row `r`.
- `dptr->get_col(c)` returns a constant pointer to an array of doubles of length `nrow`, containing values at column `c`.
- `dptr->get(r, c)` returns a double at matrix entry `(r, c)`.

In all cases, `r` and `c` should be non-negative integers in `[0, nrow)` and `[0, ncol)` respectively.
Zero-based indexing is assumed for both `r` and `c`, as is standard for most C/C++ applications.

## Other matrix types

Logical and integer matrices are also supported, by including the following header files:

```
#include "logical_matrix.h"
#include "integer_matrix.h"
```

The creation function changes correspondingly, for logical matrix `lmat` and integer matrix `imat`:

```
std::shared_ptr<logical_matrix> lptr=create_logical_matrix(lmat);
std::shared_ptr<integer_matrix> iptr=create_integer_matrix(imat);
```

Here, the `get_row` and `get_col` methods return pointers to integer arrays, while `get` returns an integer.

## Important developer information 

__NOTE:__ A pointer returned by `get_row` is not guaranteed to be valid after another call to `get_row`.
The same applies for pointers returned by `get_col` after another call to `get_col`.
This is because these pointers refer to internal arrays within each instance of the `*_matrix` class, which are overwritten upon repeated calls to the above methods.

```
const double* x=dptr->get_row(0);
x[0]; // First entry of first row, okay.
const double* y=dptr->get_row(1);
x[0]; // Not guaranteed to be unchanged!
```

If the values need to be stored, copy the values from the returned pointer into your own chunk of memory.

```
const double* x=dptr->get_row(0);
std::copy(x, x+dptr->get_ncol(), z); // Assuming you have some pointer z.
const double* y=dptr->get_row(1);
z[0]; // First entry of first row, okay.
```

For this reason, the current API is unlikely to be safe if there are multiple shared pointers to the same `*_matrix` instance (e.g., with multiple threads).
In such cases, users should instead create new shared pointers by copying the `*_matrix` instance via `std::make_shared`.

__NOTE:__ The API will happily throw exceptions of the `std::exception` class, containing an informative error message.
These should be caught and handled gracefully by the end-user code, otherwise a segmentation fault will probably occur.
See the error-handling mechanism in `r CRANpkg("Rcpp")` for an example of how to deal with these exceptions.

# Supported matrix types

The following matrix classes are supported:

- numeric: `matrix`, `dgeMatrix`, `dgCMatrix`, `dspMatrix`, `HDF5Matrix`
- integer: `matrix`, `HDF5Matrix`
- logical: `matrix`, `lgeMatrix`, `lgCMatrix`, `lspMatrix`, `HDF5Matrix`

Additional classes can be added on a need-to-use basis.
As a general rule, if a matrix-like object can be stored in a `SummarizedExperiment` class (from the `r Biocpkg("SummarizedExperiment")` package), the API should be able to handle it.
Please contact the maintainers if you have a class that you would like to see supported.

# Session information

```{r}
sessionInfo()
```
