<!--
%\VignetteIndexEntry{A C++ API for R matrix objects}
%\VignettePackage{BiocStyle}
%\VignetteEngine{knitr::knitr}
-->

```{r, echo=FALSE, results="hide", message=FALSE}
require(knitr)
opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

```{r style, echo=FALSE, results='asis'}
BiocStyle::markdown()
```

# A common C++ API for all types of R matrices
Package: `r Rpackage("beachmat")` <br />
Author: Aaron Lun (alun@wehi.edu.au) <br />
Compilation date: `r Sys.Date()`

# Introduction

This package provides a C++ API for handling a variety of R matrix types.
The aim is to abstract the specific type of matrix when writing C++ extensions to R packages, thus simplifying the processing of data stored in matrices or matrix-like objects.
Currently, the API supports double-precision, integer and logical matrices.
Supported classes include base `matrix` objects, a number of classes from the `r CRANpkg("Matrix")` package, and disk-backed matrices from the `r CRANpkg("HDF5Array")` package.

# Overview of the API

## Linking to the package

We will assume that readers have a working knowledge of R's C-level API - see https://cran.r-project.org/doc/manuals/r-release/R-exts.html for details.
We also assume that readers are familiar with C++.

<!--
Need to add some linking instructions here.
-->

## An example with a numeric matrix

We demonstrate the use of the API for  numeric matrices.
First, we include the relevant header file (in this case, `numeric_matrix.h`):

```
#include "numeric_matrix.h"
```

A double-precision matrix object `dmat` is handled in C++ by passing the `SEXP` struct from `.Call` to `create_numeric_matrix`:

```
std::shared_dptr<numeric_matrix> dptr=create_numeric_matrix(dmat);
```

This creates a shared pointer that points to an object of the `numeric_matrix` virtual class.
The exact class depends on the type of matrix in `dmat`, though the behaviour of the user-level functions are not affected by this detail.
(Note that this requires C++11 to work.)

The available methods for this object are:

- `dptr->get_nrow()` returns an integer number of rows (`nrow`).
- `dptr->get_ncol()` returns an integer number of columns (`ncol`).
- `dptr->get_row(r)` returns a constant pointer to an array of doubles of length `ncol`, containing values at row `r`.
- `dptr->get_col(c)` returns a constant pointer to an array of doubles of length `nrow`, containing values at column `c`.
- `dptr->get(r, c)` returns a double at matrix entry `(r, c)`.

In all cases, `r` and `c` should be non-negative integers in `[0, nrow)` and `[0, ncol)` respectively.
Zero-based indexing is assumed for both `r` and `c`, as is standard for most C/C++ applications.

## Other matrix types

Logical and integer matrices are also supported, by including the following header files:

```
#include "logical_matrix.h"
#include "integer_matrix.h"
```

The creation function changes correspondingly, for logical matrix `lmat` and integer matrix `imat`:

```
std::shared_ptr<logical_matrix> lptr=create_logical_matrix(lmat);
std::shared_ptr<integer_matrix> iptr=create_integer_matrix(imat);
```

Here, the `get_row` and `get_col` methods return pointers to integer arrays, while `get` returns an integer.

## Important developer information 

__NOTE:__ A pointer returned by `get_row` is not guaranteed to be valid after another call to `get_row`.
The same applies for pointers returned by `get_col` after another call to `get_col`.
This is because these pointers refer to internal arrays within each instance of the `*_matrix` class, which are overwritten upon repeated calls to the above methods.

```
const double* x=dptr->get_row(0);
x[0]; // First entry of first row, okay.
const double* y=dptr->get_row(1);
x[0]; // Not guaranteed to be unchanged!
```

If the values need to be stored, copy the values from the returned pointer into your own chunk of memory.

```
const double* x=dptr->get_row(0);
std::copy(x, x+dptr->get_ncol(), z); // Assuming you have some pointer z.
const double* y=dptr->get_row(1);
z[0]; // First entry of first row, okay.
```

For this reason, the current API is unlikely to be safe if there are multiple shared pointers to the same `*_matrix` instance (e.g., with multiple threads).
In such cases, users should instead create new shared pointers by copying the `*_matrix` instance via `std::make_shared`.

__NOTE:__ The API will happily throw exceptions of the `std::exception` class, containing an informative error message.
These should be caught and handled gracefully by the end-user code, otherwise a segmentation fault will probably occur.
See the error-handling mechanism in `r CRANpkg("Rcpp")` for an example of how to deal with these exceptions.

# Supported matrix types

The following matrix classes are supported:

- numeric: `matrix`, `dgeMatrix`, `dgCMatrix`, `dspMatrix`, `HDF5Matrix`
- integer: `matrix`, `HDF5Matrix`
- logical: `matrix`, `lgeMatrix`, `lgCMatrix`, `lspMatrix`, `HDF5Matrix`

Additional classes can be added on a need-to-use basis.
As a general rule, if a matrix-like object can be stored in a `SummarizedExperiment` class (from the `r Biocpkg("SummarizedExperiment")` package), the API should be able to handle it.
Please contact the maintainers if you have a class that you would like to see supported.

# Session information

```{r}
sessionInfo()
```
